---
title: "{gtfsrealtime}: a Package for Working with GTFS-Realtime Transit Data in R"
subtitle: "Under review at _Findings_"
authors:
  - name: Matthew Wigginton Bhagat-Conway
    affiliations:
      - name: University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
        address: "223 E Cameron Ave, CB #3140"
        city: Chapel Hill
        state: NC
        country: USA
        postal-code: 27599
    orcid: 0000-0002-1210-2982
    email: mwbc@unc.edu
  - name: Matthew D. Palm
    affiliations:
      - name: University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
        address: "223 E Cameron Ave, CB #3140"
        city: Chapel Hill
        state: NC
        country: USA
        postal-code: 27599
    orcid: 0000-0002-8800-2777
    email: palmmatt@unc.edu
bibliography: bibliography.bib
format:
  wordcount-pdf:
    count-code-cells: false
include-in-header:
  text: |
    \usepackage{placeins}
abstract: |
  Many transit agencies publish real-time data on vehicle positions, arrival predictions, and service alerts in GTFS-realtime format. While these data provide a wealth of information, to date they have been underutilized in research. In part, this is due to a lack of tooling for reading the bespoke binary format into commonly-used research platforms. This article introduces the `{gtfsrealtime}` R package which allows reading GTFS-realtime into the R statistical programming environment.

  Keywords: gtfs, gtfsrealtime, transit, r
---

## Questions

The General Transit Feed Specification (GTFS) is one of the great successes of open data standards. GTFS is a standardized format for communicating transit schedule data [@voulgaris_predictors_2023]. While GTFS was originally intended to support customer-facing trip planning apps [@mchugh_pioneering_2013], it has seen extensive use in research [@ho_general_2021]. Many open source tools are available for analyzing GTFS data [e.g., @herszenhut_gtfstools_2022; @pereira_r5r_2021; @reynolds_social_2025; @kocsis_gtfs2net_2025], many using the R statistical programming language [@r_2026].

Originally, GTFS only provided data on scheduled arrival times. In 2011, GTFS-realtime was released, allowing agencies to provide real-time information [@rychev_introducing_2011]. To date, most analytical applications have used the original GTFS specification (retroactively renamed GTFS-static), with some exceptions [e.g., @liu_realizable_2023; @javanmard_using_2025]; a Scopus search for GTFS-realtime returns only 17 articles.^[The exact search used was `( gtfs realtime ) OR ( gtfsrealtime ) OR ( gtfs-realtime ) OR ( gtfs AND realtime )`] One factor is the lack of tooling to read GTFS-realtime's bespoke binary format. This article introduces the `{gtfsrealtime}` R package, which reads GTFS-realtime feeds into standard R data frames. It answers in the affirmative the question of whether R can be efficiently used to analyze large GTFS-realtime feeds.

<!--# while Findings means this as "what are the methods used" i.e. the types of analysis, I am taking it literally as the methods/functions exported. -->
## Methods

GTFS-realtime feeds come in three types: vehicle positions, trip updates, and service alerts. The `{gtfsrealtime}` package exposes three functions for reading these: `read_gtfsrt_positions`, `read_gtfsrt_trip_updates`, and `read_gtfsrt_alerts`, which are detailed below. There are several experimental extensions to GTFS-realtime, which are not currently supported.

`{gtfsrealtime}` can be installed through the Comprehensive R Archive Network (CRAN):

```{r}
#| eval: false
install.packages("gtfsrealtime")
```

The current version is 0.2.1. `{gtfsrealtime}` source code and development versions are available from <https://github.com/mattwigway/gtfsrealtime-r>.

For performance, `{gtfsrealtime}` has components written in Rust [@rust188]; CRAN provides binary packages for the most recent version of R on Windows and macOS (if you experience installation issues, ensure you are using the latest version of R). 

Once installed, it can be loaded (along with a few other packages we will be using):

```{r}
#| output: false
library(gtfsrealtime)
library(ggplot2) # for plotting
library(dplyr) # for general data manipulation
library(data.table) # for high-performance data filtering
library(sf) # for spatial data analysis
```

```{r}
#| include: false
library(ggunc)
```

The following sections demonstrate the main functions of the package, using example data shipped with the package. Note that the example vehicle positions are from a different time period than the other datasets and cannot be joined with them.

### `read_gtfsrt_positions`

Read vehicle positions with `read_gtfsrt_positions`. The feed used below comes from the New York City bus network. This file is included with `{gtfsrealtime}` and is compressed with `bzip2`. `{gtfsrealtime}` can read uncompressed files and files compressed with `zip`, `gzip`, or `bzip2`, from the local machine or a URL. To use with your own feed, replace `VEHICLE_POSITIONS_FILE` with your own file or URL. 

GTFS-realtime does not include timezone information. We specify a timezone so that times are automatically converted to local time. Timezones are specified in standardized TZ database format (generally `Continent/City`; [see list here](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones)). `{sf}` is an R package for geographic data [@sf1; @sf2]; specifying `as_sf=TRUE` will convert the data to `{sf}` format so it can be mapped.

```{r}
VEHICLE_POSITIONS_FILE = system.file("nyc-vehicle-positions.pb.bz2",
  package = "gtfsrealtime")

positions = read_gtfsrt_positions(
  VEHICLE_POSITIONS_FILE,
  "America/New_York",
  as_sf = TRUE
)
```

\FloatBarrier

This produces a data frame of the positions file. A few rows are shown in @tbl-pos (only the most relevant columns are shown; all are described in the documentation).

```{r}
#| echo: false
#| tbl-cap: Key columns in a vehicle position feed
#| label: tbl-pos
shorten = function (text, length) {
  lens = stringr::str_length(text)

  lenstart = floor(length / 2)
  lenend = length - 1 - lenstart

  ifelse(lens > length, paste0(
    stringr::str_sub(text, 1, lenstart),
    "\u2026",
    stringr::str_sub(text, lens - lenend + 1)
    ), text)
}

head(positions) |>
  sf::st_drop_geometry() |>
  select(vehicle_id, timestamp, trip_id, route_id, stop_id) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 13))) |>
  gt::gt() |>
    gt::fmt_datetime(timestamp)
```

\FloatBarrier

Since we read the data using `as_sf=TRUE`, we can also map the locations of all vehicles, colored by how full they are (@fig-pos). One could use other packages such as `ggspatial` [@ggspatial] to add a basemap, scale bar, etc.; this is left to the reader for brevity.


```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-cap: Bus positions in NYC
#| fig-alt: Map showing bus positions in New York City with no background; the positions create an outline of the city. They are colored by how full they are, with most having many seats available.
#| label: fig-pos
positions |>
  ggplot(aes(color = occupancy_status)) +
  geom_sf(size = 0.5) +
  theme_void()
```


### `read_gtfsrt_trip_updates`

`{gtfsrealtime}` likewise includes an example trip updates feed from New York MTA buses, which we can read with `read_gtfsrt_trip_updates`. Modify `TRIP_UPDATES_FILE` to work with your own file or URL.

```{r read-updates}
#| warning: false
TRIP_UPDATES_FILE = system.file("nyc-trip-updates.pb.bz2",
  package = "gtfsrealtime")

updates = read_gtfsrt_trip_updates(
  TRIP_UPDATES_FILE,
  "America/New_York"
)
```

\FloatBarrier

GTFS-realtime trip updates are hierarchical; one trip update can contain information about the trip as a whole as well as updates for multiple stops along that trip. `read_gtfsrt_trip_updates()` flattens that structure into a data frame. As a result, the same `id` may appear in multiple rows when the feed contains stop-level updates for multiple stops. In @tbl-arrival, we inspect all rows associated with one `id` (and show the most relevant columns). If a trip update has no stop time updates, it will appear as a single row with all the stop-specific fields NA.

```{r}
#| echo: false
#| tbl-cap: Example arrival time predictions for a single trip.
#| label: tbl-arrival
updates |>
  filter(id == first(id)) |>
  select(id, trip_id, route_id, stop_id, arrival_time) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 13))) |>
  gt::gt() |>
      gt::fmt_datetime(arrival_time)
```

When reading this example feed, `{gtfsrealtime}` warns that some GTFS-realtime entity IDs are duplicated, which is problematic since these are used to identify the rows belonging to a single trip update. In these cases, the package appends suffixes such as `_duplicated_1` so that each trip update can be represented with a unique `id` within a feed (when reading a ZIP file containing multiple feeds, `id`s are only unique within each of the contained feeds, which can be disambiguated with the `file_index` field). This deduplication applies only to the `id` field. Other fields, for example `trip_id`, are not modified so they will still correctly join with GTFS-static. Deduplication also happens with alerts and vehicle position `id`s.

\FloatBarrier

### `read_gtfsrt_alerts`

```{r read-alerts}
ALERTS_FILE = system.file("nyc-service-alerts.pb.bz2",
  package = "gtfsrealtime")

alerts = read_gtfsrt_alerts(
  ALERTS_FILE,
  "America/New_York"
)
```

\FloatBarrier

Like trip updates, alerts are hierarchical; one alert can include multiple time periods, "informed entities" (such as routes or stops), and languages. `read_gtfsrt_alerts()` flattens this structure, repeating the alert for every combination of time period, entity, and language. Therefore, the same alert `id` may appear in multiple rows.

In @tbl-alert, we select the first alert and display the route or stop fields and the alert text. This alert appears as two rows because it affects two routes.

```{r}
#| echo: false
#| tbl-cap: Example alerts data
#| label: tbl-alert
#| output: asis
cat(alerts |>
  filter(id == first(id)) |>
  select(id, start, end, route_id, stop_id, header_text) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 13)), header_text=shorten(header_text, 52)) |>
  gt::gt() |>
      gt::fmt_datetime(start) |>
      gt::as_latex() |>
      # wrap the header_text column
      stringr::str_replace("lrrlcl", "lrrlcp{3cm}"))
      
```

### Archived data

Archived data is often useful for analytical applications; `{gtfsrealtime}` also supports reading directly from ZIP files containing multiple GTFS-realtime feeds (for example, all vehicle positions over the course of a day). Examples of working with archived data, including vehicle position animations and stringline charts, are in the `{gtfsrealtime}` documentation.

## Findings {#sec-findings}

This article has introduced the `{gtfsrealtime}` R package, an open-source (MIT-licensed) package for reading GTFS-realtime data into R data frames. As far as we know, this is the only actively-maintained package for reading GTFS-realtime into R. It reads the hierarchical GTFS-realtime format into a tabular format, which is conducive to efficient analysis in R. It aims to be as user-friendly as possible, handling duplicate IDs, adding factor columns rather than numeric codes for categorical variables, and processing timestamps into R date-time objects. It is fast, reading a full day of New York City vehicle positions (_n_=2.8 million) in 16 seconds on an M1 Macbook Pro—fast enough to readily analyze months or years of data.

We are aware of only one other actively-maintained option for reading GTFS-realtime data into R. GTFS-realtime is based on the Protocol Buffers format, and as such the `{RProtoBuf}` package [@eddelbuettel_rprotobuf_2016], a general-purpose package for reading any Protocol Buffers-based format, can read it.  `{RProtoBuf}` actually reads GTFS-realtime somewhat faster than `{gtfsrealtime}` (the same full day of vehicle positions takes only 3 seconds to read). However, it returns data in a hierarchical format that is not conducive to efficient analysis in R. Converting to the tabular format necessary for further analysis takes an additional 25 minutes using `vapply` and `lapply`.^[Full details of this test are [in the Performance article available in the `{gtfsrealtime}` documentation](https://projects.indicatrix.org/gtfsrealtime-r/articles/performance.html).] `RProtoBuf` also does not provide GTFS-realtime specific features such as data compression, ID deduplication, or time zone parsing. Another package, `{gtfsway}`, has not been maintained in many years and is formally archived [@_gtfsway_2025]. There are of course also GTFS-realtime bindings in many other languages, such as Python.

There is a wealth of applications of real-time transit data---for example, accessibility analysis [e.g., @wessel_accuracy_2019], operations [e.g., @aemmer_measurement_2022], or passenger information evaluation [e.g, @newmark_assessing_2024]. We hope that the `{gtfsrealtime}` package will support these types of analysis and more, and will underly development of additional standardized packages for common analyses using realtime data.

## References