---
title: "{gtfsrealtime}: a Package for Working with GTFS-Realtime Transit Data in R"
subtitle: "Word count: {{< words-body >}}"
authors:
  - name: Matthew Wigginton Bhagat-Conway
    affiliations:
      - name: University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
        address: "223 E Cameron Ave, CB #3140"
        city: Chapel Hill
        state: NC
        country: USA
        postal-code: 27599
    orcid: 0000-0002-1210-2982
    email: mwbc@unc.edu
  - name: Matthew D. Palm
    affiliations:
      - name: University of North Carolina at Chapel Hill
        department: Department of City and Regional Planning
        address: "223 E Cameron Ave, CB #3140"
        city: Chapel Hill
        state: NC
        country: USA
        postal-code: 27599
    orcid: 0000-0002-8800-2777
    email: palmmatt@unc.edu
bibliography: bibliography.bib
format:
  wordcount-pdf:
    count-code-cells: false
include-in-header:
  text: |
    \usepackage{placeins}
---

```{r}
#| echo: false
# make sure the GTFS archived data is available
suppressMessages({
  if (!file.exists("nyc-bus-demo.zip")) {
    download.file(
      "https://files.indicatrix.org/gtfsrealtime-r/nyc-bus-demo.zip",
      "nyc-bus-demo.zip"
    )
  }

    if (!file.exists("brooklyn-bus.zip")) {
    download.file(
      "https://files.indicatrix.org/gtfsrealtime-r/brooklyn-bus.zip",
      "brooklyn-bus.zip"
    )
  }
})
```

## Questions

The General Transit Feed Specification (GTFS) is one of the great successes of open data standards. GTFS is a standardized format for communicating transit schedule data. First released in 2006, by 2022 it had achieved a market penetration of over 75% in the US [@voulgaris_predictors_2023]. While GTFS was originally intended for sharing data with developers of customer-facing transit apps such as trip planners [@mchugh_pioneering_2013], it has also seen extensive use in research [@ho_general_2021]. There are many open source tools available for analyzing GTFS data [e.g., @herszenhut_gtfstools_2022; @pereira_r5r_2021; @pereira_exploring_2023; @reynolds_social_2025; @kocsis_gtfs2net_2025], many implemented in the R statistical programming language [@r_2026].

The original GTFS specification only provides data on scheduled arrival times. In 2011, GTFS-realtime was released (and the original specification retroactively renamed GTFS-static), allowing agencies to provide real-time arrival time predictions, vehicle positions, and service alerts [@gtfrst_2026]. To date, most analytical applications have used GTFS-static, with some exceptions [e.g., @lee_robust_2020; @liu_realizable_2023; @javanmard_using_2025]. We hypothesize there are two reasons for this gap. First, GTFS-realtime data is ephemeral, designed to drive customer-facing apps in real time. To use it in research, one must archive the data at high frequency (minutely seems standard) for the entire analysis period. Second, there is a lack of standardized tooling for working with GTFS-realtime data, with most tools focusing on GTFS-static data. Furthermore, the GTFS-realtime format is significantly more complicated than GTFS-static. GTFS-static is a ZIP file containing CSV files, which can easily be read into most data-manipulation platforms without specialized tooling (though specialized tooling certainly makes working with it more pleasant). On the other hand, GTFS-realtime is a custom binary format based on Protocol Buffers [@jean_protocol_2024] which cannot be read without specialized tooling.

This article introduces the `{gtfsrealtime}` R package, which efficiently reads GTFS-realtime feeds into standard R data frames for further analysis.

<!--# while Findings means this as "what are the methods used" i.e. the types of analysis, I am taking it literally as the methods/functions exported. -->
## Methods

GTFS-realtime feeds come in three types: vehicle positions, trip updates, and service alerts. While it is legal to mix these into a single file, generally agencies provide three separate files for the different types. The `{gtfsrealtime}` package exposes three functions for reading each of these types of files: `read_gtfsrt_positions`, `read_gtfsrt_trip_updates`, and `read_gtfsrt_alerts`, which are detailed below. In the rare case where the types are combined into a single file, each function will read the type of data requested and ignore other types of data. 

`{gtfsrealtime}` is available through the Comprehensive R Archive Network (CRAN), and can be installed in the usual way:

```{r}
#| eval: false
install.packages("gtfsrealtime")
```

For performance, `{gtfsrealtime}` has compiled components written in the Rust programming language [@klabnik_rust_2026]. CRAN provides pre-compiled binary packages for the most recent version of R on Windows and macOS; we recommend users use these. (If you get error messages about `rustc` not being available, you likely need to upgrade to the latest version of R so you can use the pre-compiled packages). For Linux users, [the Posit Public Package Manager](https://packagemanager.posit.co/) provides compiled binary packages.

`{gtfsrealtime}` is open-source; source code and development versions are available from <https://github.com/mattwigway/gtfsrealtime-r>.

Once installed, it can be loaded (along with a few other packages we will be using):

```{r}
#| output: false
library(gtfsrealtime)
library(ggplot2) # for plotting
library(dplyr) # for general data manipulation
library(data.table) # for high-performance data filtering
library(sf) # for spatial data analysis
```

```{r}
#| include: false
library(ggunc)
```

### `read_gtfsrt_positions`

Vehicle position feeds can be read with `read_gtfsrt_positions`. The feed used below comes from the New York City bus network. This file is included with {gtfsrealtime} and is compressed with `bzip2` to save space. All functions in {gtfsrealtime} can automatically detect and read uncompressed files as well as files compressed with `zip`, `gzip`, or `bzip2`. It can also read directly from an `http://` or `https://` URL. To use with your own feed, simply replace `VEHICLE_POSITIONS_FILE` with the path to your own file.

GTFS-realtime does not include time zone information; all times are specified in UTC. We specify a time zone so that times are automatically converted to local time. Time zones are specified in standardized TZ database format (generally `Continent/City`; for a list, [see here](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones)). If you do not want to convert times, you can specify a time zone of `Etc/UTC`. The `{sf}` package is an R package for geographic data manipulation [@sf1; @sf2]; specifying `as_sf=TRUE` will convert the data to `{sf}` format so it can be mapped or analyzed spatially.

```{r}
VEHICLE_POSITIONS_FILE = system.file("nyc-vehicle-positions.pb.bz2",
  package = "gtfsrealtime")

positions = read_gtfsrt_positions(
  VEHICLE_POSITIONS_FILE,
  "America/New_York",
  as_sf = TRUE
)
```

\FloatBarrier

This produces a data frame with the information from the GTFS-realtime positions file. The first few rows are shown in @tbl-pos (with only the most relevant columns shown to save space; all columns are fully documented in the documentation that ships with `{gtfsrealtime}`).

```{r}
#| echo: false
#| tbl-cap: Key columns in a vehicle position feed
#| label: tbl-pos
shorten = function (text, length) {
  lens = stringr::str_length(text)

  lenstart = floor(length / 2)
  lenend = length - 1 - lenstart

  ifelse(lens > length, paste0(
    stringr::str_sub(text, 1, lenstart),
    "\u2026",
    stringr::str_sub(text, lens - lenend + 1)
    ), text)
}

head(positions) |>
  sf::st_drop_geometry() |>
  select(vehicle_id, timestamp, trip_id, route_id, stop_id) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 12))) |>
  gt::gt() |>
    gt::fmt_datetime(timestamp)
```

Since we read the data using `as_sf = TRUE`, we can also map the locations of all vehicles in the feed, colored by how full they are (@fig-pos).

```{r}
#| fig-width: 6
#| fig-height: 4
#| fig-cap: Bus positions in NYC
#| fig-alt: Map showing bus positions in New York City with no background; the positions create an outline of the city. They are colored by how full they are, with most having many seats available.
#| label: fig-pos
positions |>
  ggplot(aes(color = occupancy_status)) +
  geom_sf(size = 0.1) +
  theme_void()
```


### `read_gtfsrt_trip_updates`

`{gtfsrealtime}` likewise includes an example trip updates feed from New York MTA buses, which we can read with `read_gtfsrt_trip_updates`. As before, you can work with your own feed by replacing `TRIP_UPDATES_FILE` with your own file or URL.

```{r read-updates}
#| warning: false
TRIP_UPDATES_FILE = system.file("nyc-trip-updates.pb.bz2",
  package = "gtfsrealtime")

updates = read_gtfsrt_trip_updates(
  TRIP_UPDATES_FILE,
  "America/New_York"
)
```

\FloatBarrier

When reading this example feed, `{gtfsrealtime}` warns that some GTFS-realtime entity IDs are duplicated. In these cases, the package appends suffixes such as `_duplicated_1` so that each row can be represented with a unique `id`. There are quite a few of them, so they are suppressed for readability, but the first two are:

```
1: ! ID UP_A6-Weekday-SDon-094800_B6_243 is duplicated. Replacing with UP_A6-Weekday-SDon-094800_B6_243_duplicated_1 . This may cause joins between different
  GTFS-realtime files (even within a ZIP archive) to be incorrect.
2: ! ID UP_A6-Weekday-SDon-094800_B6_243 is duplicated. Replacing with UP_A6-Weekday-SDon-094800_B6_243_duplicated_2 . This may cause joins between different
  GTFS-realtime files (even within a ZIP archive) to be incorrect.
```

These warnings are useful in practice: duplicated entity IDs can affect workflows that join records across GTFS-realtime files, or with GTFS-static.

GTFS-realtime trip updates are [hierarchical](https://gtfs.org/documentation/realtime/reference/#message-tripupdate); one trip update can contain information about the trip as a whole as well as updates for multiple stops along that trip. `read_gtfsrt_trip_updates()` flattens that structure into a data frame. As a result, the same `id` may appear in multiple rows when the feed contains stop-level updates for multiple stops. In @tbl-arrival, we inspect all rows associated with one `id` (and show the most relevant columns). If a trip update has no stop time updates, it will appear as a single row with all the stop-specific fields NA.

```{r}
#| echo: false
#| tbl-cap: Example arrival time predictions for a single trip.
#| label: tbl-arrival
updates |>
  filter(id == first(id)) |>
  select(id, trip_id, route_id, stop_id, arrival_time) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 12))) |>
  gt::gt() |>
      gt::fmt_datetime(arrival_time)
```

### `read_gtfsrt_alerts`


```{r read-alerts}
ALERTS_FILE = system.file("nyc-service-alerts.pb.bz2",
  package = "gtfsrealtime")

alerts = read_gtfsrt_alerts(
  ALERTS_FILE,
  "America/New_York"
)
```

\FloatBarrier

Like trip updates, GTFS-realtime alerts are [hierarchical](https://gtfs.org/documentation/realtime/reference/#message-alert): one alert can include multiple active time periods, multiple "informed entities" (such as routes, stops, trips, or agencies), and multiple languages affected by the alert. `read_gtfsrt_alerts()` flattens this structure into a data frame, repeating the alert for every combination of time period, entity, and language. As a result, the same alert `id` may appear in multiple rows when an alert applies to more than one entity or time period.

Because a single alert can apply to more than one route, stop, or trip, it is useful to inspect all rows associated with one alert id. In @tbl-alert, we select the first alert in the feed and display the affected route or stop fields along with the rider-facing alert text. This alert appears as two separate rows because it affects two routes: the X28 and the X38.

```{r}
#| echo: false
#| tbl-cap: Example alerts data
#| label: tbl-alert
#| output: asis
cat(alerts |>
  filter(id == first(id)) |>
  select(id, start, end, route_id, stop_id, header_text) |>
  mutate(across(ends_with("id"), \(x) shorten(x, 12))) |>
  gt::gt() |>
      gt::fmt_datetime(start) |>
      gt::as_latex() |>
      # wrap the header_text column
      stringr::str_replace("lrrlcl", "lrrlcp{3cm}"))
      
```

### Archived data

A GTFS-realtime feed is a snapshot in time of a transit system. Most often, it will be more analytically interesting to work with a group of feeds archived over some time period. Frequent archiving can quickly create an unwieldy number of files, so all `{gtfsrealtime}` functions can work with a ZIP file containing multiple feeds. For instance, the code below will read all bus positions from the New York City MTA over the course of a day, from a ZIP file containing one feed per minute, and converts it to a `data.table` for efficient manipulation [@data.table]. This file is too large to include with the package, but is available for download at <https://files.indicatrix.org/gtfsrealtime-r/nyc-bus-demo.zip>. Reading the whole day of vehicle locations takes 44 seconds on an M1 Macbook Pro.

```{r}
daily_positions = read_gtfsrt_positions("nyc-bus-demo.zip", "America/New_York") |>
  data.table()
```

Some of the observations are duplicates, as not every bus reports in every minute. The syntax below groups by timestamp and vehicle ID and then selects a single observation for each. At least in the New York data, there are sometimes multiple reports with the same timestamp and vehicle ID but different locations. This is a data error, so we arbitrarily pick one. After deduplication, we have 2.8 million observations.

```{r}
daily_positions = daily_positions[, .SD[1], by = c("timestamp", "vehicle_id")]
nrow(daily_positions)
```

Since there are multiple feeds within the ZIP file, the `file_index` field will contain a numeric identifier for each feed. Note that this is based on the order the feeds occur in the ZIP file, it may not be chronological. There is also a `file_timestamp` field identifying when the file containing each observation was produced, but this is not guaranteed to be unique across files.

For example, by combining a file like this with the static GTFS, we can make a stringline plot showing the progress of vehicles in each direction along the route. The plot clearly shows that there was some bunching on this day around rush hour (@fig-string). The code to produce this plot is not reproduced here for brevity but is in the "Stringline" vignette included with the package.

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 6
#| fig-cap: Stringline plot of vehicles on New York City MTA route B4.
#| fig-alt: Plot showing time on the x axis and stops on the y axis, with diagonal lines indicating vehicle progress along the route. Buses bound for Knapp St/Harkness Ave are in pink, while buses bound for Narrows Ave/77 St afe in green. In a number of places there are large gaps along the x axis followed by many lines of one color, indicating bunching.
#| label: fig-string
# find all trips on the B4 with direction id 0
static = gtfstools::read_gtfs("brooklyn-bus.zip")

ROUTE_ID = "B4"
trips = static$trips[route_id == ROUTE_ID & direction_id == 0, trip_id]
stop_times = static$stop_times[trips, on = .(trip_id)]
fwd_stop_sequences = stop_times[, .(stop_sequence = median(stop_sequence), timepoint = first(timepoint)), by = stop_id]
fwd_stop_sequences = static$stops[fwd_stop_sequences, on = .(stop_id)]

# now, figure out the reverse direction
reverse_trips = static$trips[route_id == ROUTE_ID & direction_id == 1, trip_id]
reverse_stop_times = static$stop_times[reverse_trips, on = .(trip_id)]
reverse_stops = unique(reverse_stop_times[, .(stop_id)])
reverse_stops = static$stops[reverse_stops, on = .(stop_id)]

# spatially join the reverse stops to the forward stops to get the closest stop sequence
fwd_stop_sequences = st_as_sf(fwd_stop_sequences, coords = c("stop_lon", "stop_lat"), crs = 4326) |>
  st_transform(32118)
reverse_stops = st_as_sf(reverse_stops, coords = c("stop_lon", "stop_lat"), crs = 4326) |>
  st_transform(32118)
reverse_stops = st_join(reverse_stops, select(fwd_stop_sequences, stop_sequence), join = st_nearest_feature)

stop_sequences = rbind(
  select(fwd_stop_sequences, stop_id, stop_sequence) |> st_drop_geometry(),
  select(reverse_stops, stop_id, stop_sequence) |> st_drop_geometry()
) |>
  # make sure stops that appear in both aren't duplicated
  group_by(stop_id) |>
  slice_head(n = 1) |>
  ungroup()

route_positions = daily_positions |>
  filter(route_id == ROUTE_ID)

route_positions |>
  left_join(stop_sequences, by = "stop_id") |>
  left_join(select(fwd_stop_sequences, stop_sequence, stop_name), by = "stop_sequence") |>
  group_by(vehicle_id, trip_id, stop_sequence) |>
  slice_max(timestamp) |> # the latest time reported before passing the stop
  ungroup() |>
  arrange(vehicle_id, timestamp) |>
  ggplot(aes(
    x = reorder(stringr::str_to_title(stop_name), stop_sequence),
    y = timestamp,
    group = paste(vehicle_id, trip_id),
    color = factor(direction_id)
  )) +
  geom_path(size = 0.1) +
  theme(legend.position = "none") +
  xlab("Stop (timepoints labeled)") +
  ylab("Time") +
  scale_x_discrete(breaks = stringr::str_to_title(fwd_stop_sequences[fwd_stop_sequences$timepoint == 1, ]$stop_name)) +
  coord_flip() +
  scale_y_datetime(labels = \(dt) strftime(dt, "%H:%M"))
```

## Findings {#sec-findings}

This article has introduced the `{gtfsrealtime}` R package, an efficient package for reading GTFS-realtime. There is a wealth of applications of real-time transit data---for example, accessibility analysis [e.g., @wessel_accuracy_2019], operations [e.g., @aemmer_measurement_2022], or passenger information evaluation [e.g, @newmark_assessing_2024]. We hope that the `{gtfsrealtime}` package will support these types of analysis and more, and will underly development of additional standardized packages for common analyses using realtime data.